---
title: Minicurso Multi-trait <br> LAPEA
#subtitle: 
author: "Mateus Gonçalves"
bibliography: references.bib
date: "03/17/2025"
date-format: "DD/MM/YYYY"
format:
  revealjs:
    width: 1600
    height: 800
    title-slide-attributes:
      data-background-size: contain
      class: "no-footer-slide"  # Add a custom class to the title slide
    css: styles.css
    logo: Images\lapea_logo.jpeg
   # footer: "Code available on #[GitHub](https://github.com/goncalvesmateu/Multi-trait-LAPEA.git)."
    self-contained: true
    scrollable: false
    theme: default # built-in themes for reveal presentations
    background-transition: none
    progress: false # Show a progress bar at the bottom (defaults to true).
    history: true
    toc: false
    transition: slide
    transition-speed: default
    highlight-style: github # specify the code highlighting style
    code-line-numbers: true
    code-copy: hover
    controls: auto # navigation arrows per slide
    touch: true
    code-fold: true
    multiplex: false
    chalkboard: false
    slide-level: 2
    scroll-view: true
    slide-number: true
    show-slide-number: all
    callout-appearance: simple
    preview-links: auto
    incremental: false
editor: visual
---

## Outline{style="color: #081539"}

::: {style="font-size: 75%;"}
### [Background]{style="color:black"}

- [Phenotypic, genetic and environmental correlation]

- [Direct and indirect selection]

### [Multiple trait selection] {style="color:black"}

-   [Multistage selection]
-   [Tandem selection]
-   [Independent culling levels]
-   [Index selection]

### [Multi-trait model] {style="color:black"}

- [Uni-trait model]
- [Multi-trait model]
- [Hands-on using R]
:::

## Background 

::: {layout="[ 0.8, -0.3, 0.5 ]"}

-   Breeders select for multiple traits (morphological, physiological, disease resistance...)

-   The list of traits can be endless
Depending on the program or project goals, one of three types of multiple-trait selection can be employed. A brief description of these types follows.


![**Fig 1**. *Model or ideal plant with attributes <br> combined to maximize yield* <br> [@furbank2019field].](Images\ideotypeWheat.png){fig-align="left"}
:::

## Phenotypic, genetic and environmental correlation {style="font-size: 90%;"}


## Direct and indirect selection



## Multiple trait selection {.center}

Examples of multiple-trait selection strategies can be employed... 

## Multistage selection

Multistage selection: Selection for different traits at different stages during cultivar development.

```{r fig.align='center', fig.cap=""}

set.seed(1234)
x <- rnorm(200, mean = 10, sd = 4)
y <- rnorm(200, mean = 6, sd = 2)

dat <- data.frame(x,y)

# Visualize the scatter plot of traits A and B
plot(x,y, pch=19, ylim = c(min(dat$y), max(dat$y)),xlim = c(min(dat$x), max(dat$x)),
     rect(par("usr")[1], par("usr")[3],
          par("usr")[2], par("usr")[4], col = "lightgrey"),
     cex.lab=1.5, cex.axis=1.5,
     xlab="Trait A", ylab="Trait B",col=ifelse(x > 10 & y >5, 'red', 'blue'))
abline(v=10, h=5, col="black", lwd = 2, lty = 2)

```

## Tandem selection

Selection for one trait until that trait is improved, then for a second, etc., until finally each has been improved to the desired level.

```{r fig.align='center', fig.cap=""}

set.seed(1234)
x <- rnorm(200, mean = 10, sd = 4)
y <- rnorm(200, mean = 6, sd = 2)

dat <- data.frame(x,y)

# Visualize the scatter plot of traits A and B
plot(x,y, pch=19, ylim = c(min(dat$y), max(dat$y)),xlim = c(min(dat$x), max(dat$x)),
     rect(par("usr")[1], par("usr")[3],
          par("usr")[2], par("usr")[4], col = "lightgrey"),
     cex.lab=1.5, cex.axis=1.5,
     xlab="Trait A", ylab="Trait B",col=ifelse(x > 10 & y >5, 'red', 'blue'))
abline(v=10, h=5, col="black", lwd = 2, lty = 2)

```

## Independent culling levels

A certain level of merit is established for each trait, and all individuals below that level are discarded regardless of values for other traits.

```{r fig.align='center', fig.cap=""}

set.seed(1234)
x <- rnorm(200, mean = 10, sd = 4)
y <- rnorm(200, mean = 6, sd = 2)

dat <- data.frame(x,y)

# Visualize the scatter plot of traits A and B
plot(x,y, pch=19, ylim = c(min(dat$y), max(dat$y)),xlim = c(min(dat$x), max(dat$x)),
     rect(par("usr")[1], par("usr")[3],
          par("usr")[2], par("usr")[4], col = "lightgrey"),
     cex.lab=1.5, cex.axis=1.5,
     xlab="Trait A", ylab="Trait B",col=ifelse(x > 10 & y >5, 'red', 'blue'))
abline(v=10, h=5, col="black", lwd = 2, lty = 2)

```

## Index selection

Select for all traits simultaneously by using some index of net merit.

```{r fig.align='center', fig.cap=""}

set.seed(1234)
x <- rnorm(200, mean = 10, sd = 4)
y <- rnorm(200, mean = 6, sd = 2)

dat <- data.frame(x,y)

# Visualize the scatter plot of traits A and B
plot(x,y, pch=19, ylim = c(min(dat$y), max(dat$y)),xlim = c(min(dat$x), max(dat$x)),
     rect(par("usr")[1], par("usr")[3],
          par("usr")[2], par("usr")[4], col = "lightgrey"),
     cex.lab=1.5, cex.axis=1.5,
     xlab="Trait A", ylab="Trait B",col=ifelse(x > 10 & y >5, 'red', 'blue'))
abline(v=10, h=5, col="black", lwd = 2, lty = 2)

```

## Another slide 3

::: columns
::: {.column width="40%"}
``` tex
\begin{gather*}
a_1=b_1+c_1\\
a_2=b_2+c_2-d_2+e_2
\end{gather*}

\begin{align}
a_{11}& =b_{11}&
  a_{12}& =b_{12}\\
a_{21}& =b_{21}&
  a_{22}& =b_{22}+c_{22}
\end{align}
```
:::

::: {.column width="60%"}
```{=tex}
\begin{gather*}
a_1=b_1+c_1\\
a_2=b_2+c_2-d_2+e_2
\end{gather*}
```
```{=tex}
\begin{align}
a_{11}& =b_{11}&
  a_{12}& =b_{12}\\
a_{21}& =b_{21}&
  a_{22}& =b_{22}+c_{22}
\end{align}
```
:::
:::

## Uni-trait model


$$
y = Xb + Zu + e
$$


## Multi-trait model


\begin{gather*}
y_{i} = X_{i}b_{i} + Z_{i}u_{i} + e_{i} , & i = 1, 2
\end{gather*}

\begin{align}

\begin{bmatrix}
   y_{1} \\
   y_{2} \\
   \end{bmatrix} = 
   
   \begin{bmatrix}
   X_{1} & 0 \\
   0 & X_{2} \\
   \end{bmatrix} \begin{bmatrix}
   b_{1} \\
   b_{2} \\
   \end{bmatrix} +
   
    \begin{bmatrix}
   Z_{1} & 0 \\
   0 & Z_{2} \\
   \end{bmatrix}\begin{bmatrix}
   u_{1} \\
   u_{2} \\
   \end{bmatrix} +
   
   \begin{bmatrix}
   e_{1} \\
   e_{2} \\
   \end{bmatrix}
   
\end{align}

## Advantages


## Hands-on using R

-   [Material on GitHub](https://github.com/goncalvesmateu/Multi-trait-LAPEA)

``` tex
 Installing packages
```

## References